Unimodular covers of 3-dimensional parallelepipeds and Cayley sums

### 2023-06-19 unimodular-covers-CombTheory-final.tex
edits to setup in Section 4

\begin{lemma}
\label{lemma:cayley}
Let $Q\subset \R^2$ be a two-dimensional lattice polytope and $P\subset \R^2$ a weak Minkowski summand of it. 
Let $p=[p_1,p_2] \subset P$ and $q=[q_1,q_2]\subset Q$ be two primitive and non-parallel lattice segments, and let $\vecline p$ and $ \vecline q$ be the lines spanned by them.  If the parallelogram $p + q$ is not unimodular, then at least one of the regions
\[
((p_1, p_2) + \vecline q ) \cap P, 
\qquad \text{and} \qquad
((q_1, q_2) + \vecline p ) \cap Q
\]
contains a lattice point. 
\end{lemma}

@@

\section{Proof of \Cref{lemma:cayley}}
\label{sec:the_lemma}
Let $f_q$ be the primitive lattice functional constant on $q$ and $f_p$ the one constant on $p$. We assume that $f_q(p_1) < f_q(p_2)$ and $f_p(q_1) < f_p(q_2)$.
Observe that since the segments $p_1p_2$ and $q_1q_2$ are primitive, the fact the $f_p$ and $f_q$ are also primitive implies that the area of the parallelogram $p+q$ equals both $f_p(q_2) - f_p(q_1)$ and $f_q(p_2) - f_q(p_1)$.
%
Since we can perform without loss of generality respective lattice translations to $P$ and to $Q$, we assume that $q_1$ is the origin, that $p$ belongs to the strip $q+\vecline p$  and that $f_q(p_1)=-1$.

Then $f_q(p_2)$ must be strictly positive, since $f_q(p_2)=0$ would imply that $p+q$ is a unimodular paralellogram. Moreover, since $p$ is primitive  $p$ must be in the interior of $q +\vecline p$ or otherwise we would have $p_2$ equal to one of $q_1$ or $q_2$, and then $f_q(p_2)=0$.
This implies that $p_1$ is the unique lattice point with $f_q(x)=-1$ in $q +\vecline p$. Similarly, the unique lattice point in the strip with $f_q(x)=1$ is  $q_1+q_2 -p_1$.

@@

Let $w=\area(p+q) \geq 2$, where $\area$ denotes the area normalized to a fundamental domain. In what follows, the width of a functional $f$ on a set $S$, denoted $\width_f(S)$ is defined as the difference $\sup_{x\in S} f(x)-\inf_{x\in S} f(x)$.
Then:
\[
w=\width_{f_q}(p + \vecline q )=\width_{f_q}(p)=\width_{f_p}(q)=\width_{f_p}(q +\vecline p ). 
\]

@@

Let $g$ be the primitive lattice functional constant on $[p_1, q_2]$ (and therefore constant also on $[q_1, q_1+q_2-p_1]$). By the assumption on $p_1$, the values of $g$ on these segments differ by $1$. We choose the sign of $g$ so that 
\[
g([p_1, q_2])= g( [q_1, q_1+q_2-p_1]) -1. 
\]

>>>>

\begin{lemma}
\label{lemma:cayley}
Let $Q\subset \R^2$ be a two-dimensional lattice polytope and $P\subset \R^2$ a weak Minkowski summand of it. 
Let $p=[p_1,p_2] \subset P$ and $q=[q_1,q_2]\subset Q$ be two primitive and non-parallel lattice segments, and let $\vecline p$ and $ \vecline q$ be the vector lines spanned by them.  If the parallelogram $p + q$ is not unimodular, then at least one of the regions
\[
((p_1, p_2) + \vecline q ) \cap P, 
\qquad \text{and} \qquad
((q_1, q_2) + \vecline p ) \cap Q
\]
contains a lattice point. 
\end{lemma}

@@

\section{Proof of \Cref{lemma:cayley}}
\label{sec:the_lemma}

Let $f_q$ be the primitive lattice (affine) functional vanishing on $q$ and $f_p$ the one vanishing on $p$. We assume that $f_q(p_1) < f_q(p_2)$ and $f_p(q_1) < f_p(q_2)$.
Let $w=\area(p+q) \geq 2$, where $\area$ denotes the area normalized to a fundamental domain. In what follows, the width of a functional $f$ on a set $S$, denoted $\width_f(S)$ is defined as the difference $\sup_{x\in S} f(x)-\inf_{x\in S} f(x)$.
Then primitiveness of $p$, $q$, $f_p$ and $f_q$ implies that:
\begin{align*}
w=&\width_{f_q}(p + \vecline q )=\width_{f_q}(p) = f_q(p_2) - f_q(p_1) =\\
=&\width_{f_p}(q +\vecline p ) = \width_{f_p}(q)= f_p(q_2) - f_p(q_1).
\end{align*}

%
Since we can perform without loss of generality respective lattice translations to $P$ and to $Q$, we assume that the segments $p$ and $q$ intersect in their relative interiors and that $f_q(p_1)=-1$. That is,
\[
-1=f_q(p_1) < 0 < f_q(p_1), \quad
-f_p(q_1) < 0 < f_p(q_2).
\]
In particular, $p$ must be in the interior of $q +\vecline p$ and $q$ in the interior of $p +\vecline q$
Our assumption also implies that $p_1$ is the unique lattice point with $f_q(x)=-1$ in $q +\vecline p$. Similarly, the unique lattice point in the strip with $f_q(x)=1$ is  $q_1+q_2 -p_1$.
See Figure~\ref{fig:setup} for an illustration of this setup.

@@

@@

Let $g$ be the primitive lattice functional constant on $[p_1, q_2]$ (and therefore constant also on $[q_1, q_1+q_2-p_1]$). By the assumption that $f_q(p_1)=-1$ the quadrilateral spanned by these two parallel segments is unimodular, so  the values of $g$ on them differ by $1$. We choose the sign of $g$ so that 
\[
g([p_1, q_2])= g( [q_1, q_1+q_2-p_1]) -1. 
\]
